---
title: "Bike Rental Prediction"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Synopsis

Bike sharing systems are a means of renting bicycles where the process of obtaining membership, rental, and bike return is automated via a network of kiosk locations throughout a city. Using these systems, people are able rent a bike from a one location and return it to a different place on an as-needed basis. Currently, there are over 500 bike-sharing programs around the world. In this project, we will use linear regression and a log transformation to predict bike sharing demand.

## Data

The data is taken from [Kaggle](https://www.kaggle.com/c/bike-sharing-demand/overview). It contains the two-year historical log corresponding to years 2011 and 2012 from Capital Bikeshare system, Washington D.C., USA. Details of its columns are as follows:

| Variable   | Description                                                                                                                                                                                                                                                                 |
|------------|-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------|
| datetime   | hourly date + timestamp                                                                                                                                                                                                                                                     |
| season     | 1 = spring, 2 = summer, 3 = fall, 4 = winter                                                                                                                                                                                                                                |
| holiday    | whether the day is considered a holiday                                                                                                                                                                                                                                     |
| workingday | whether the day is neither a weekend nor holiday                                                                                                                                                                                                                            |
| weather    | 1: Clear, Few clouds, Partly cloudy, Partly cloudy 2: Mist + Cloudy, Mist + Broken clouds, Mist + Few clouds, Mist 3: Light Snow, Light Rain + Thunderstorm + Scattered clouds, Light Rain + Scattered clouds 4: Heavy Rain + Ice Pallets + Thunderstorm + Mist, Snow + Fog |
| temp       | temperature in Celsius                                                                                                                                                                                                                                                      |
| atemp      | "feels like" temperature in Celsius                                                                                                                                                                                                                                         |
| humidity   | relative humidity                                                                                                                                                                                                                                                           |
| windspeed  | wind speed                                                                                                                                                                                                                                                                  |
| casual     | number of non-registered user rentals initiated                                                                                                                                                                                                                             |
| registered | number of registered user rentals initiated                                                                                                                                                                                                                                 |
| count      | number of total rentals                                                                                                                                                                                                                                                     |

## Loading in the Data

Load in required libraries.
```{r}
library(lubridate)
library(xts)
library(ggplot2)
library(gridExtra)
library(corrplot)
library(caret)
library(MASS)
library(dplyr)
library(Metrics)
```

Check its head and structure.
```{r}
rental <- read.csv("train.csv")
head(rental)
str(rental)
```

## Data Cleaning

Convert the following columns to categorical.
```{r}
rental$season <- factor(rental$season)
rental$holiday <- factor(rental$holiday)
rental$workingday <- factor(rental$workingday)
rental$weather <- factor(rental$weather)
```

Note that there are multiple rows with `windspeed` value of 0.
```{r}
length(rental$windspeed[rental$windspeed == 0])
```

Replace '0' values with previous known value, as wind speed at a point in time is dependent on the wind speed an hour ago. Since the first few rows have '0' wind speed and no reference point, we replace them with the next known value.
```{r}
is.na(rental$windspeed) <- rental$windspeed == 0
rental$windspeed <- na.locf(rental$windspeed, na.rm=FALSE)
# Replace first few '0' values with next known value
rental$windspeed <- na.locf(rental$windspeed, fromLast = TRUE, na.rm=FALSE)
```


## Feature Engineering

Create multiple columns from `datetime` column, including another season column created from the months. Convert the columns to categorical and drop the `datetime` column.
```{r}
rental$datetime <- ymd_hms(rental$datetime)
rental$week <- factor(week(rental$datetime))
rental$hour <- factor(hour(rental$datetime))
rental$day <- factor(day(rental$datetime))
rental$weekday <- factor(weekdays(rental$date), levels= c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
rental$month <- month(rental$datetime)
rental$season2 <- ifelse((rental$month <=5) & (rental$month >=3), 1, 
                        ifelse((rental$month <=8) & (rental$month >=6), 2,
                                ifelse((rental$month <=11) & (rental$month >=9), 3, 4)))
rental$month <- factor(rental$month)
rental$season2 <- factor(rental$season2)
rental <- subset(rental, select = -datetime)
```

When we compare the temperatures for `season` and `season2`, we realise that `season` has higher temperatures in autumn than summer, which does not seem accurate. Thus, we are going to use `season2` instead of `season`.
```{r}
box1 <- rental %>% group_by(season) %>% select(temp) %>% ggplot(aes(x=season,y=temp,fill=season)) + geom_boxplot()
box2 <- rental %>% group_by(season2) %>% select(temp) %>% ggplot(aes(x=season2,y=temp,fill=season2)) + geom_boxplot()
grid.arrange(box1, box2, nrow=2)
```

Drop the `season` column.
```{r}
rental <- subset(rental, select = -season)
```



## Exploratory Data Analysis

### Removing Outliers

Use boxplots to check for outliers.
```{r}
box3 <- ggplot(rental, aes(y=count)) + geom_boxplot(fill='tomato')
box4 <- ggplot(rental, aes(x=season2, y=count, fill=season2)) + geom_boxplot()
box5 <- ggplot(rental, aes(x=weekday, y=count, fill=weekday)) + geom_boxplot() + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
box6 <- ggplot(rental, aes(x=workingday, y=count, fill=workingday)) + geom_boxplot()
grid.arrange(box3, box4, box5, box6, ncol=2)
```

Based on the boxplot rule, there appears to be multiple outliers. We remove the outliers as follows:
```{r}
upperLimit <- quantile(rental$count, 0.75) + 1.5 * IQR(rental$count)
rentalOut <- rental %>% filter(count <= upperLimit)
```

Check the boxplots again and note that there are fewer outliers left.
```{r}
box3 <- ggplot(rentalOut, aes(y=count)) + geom_boxplot(fill='tomato')
box4 <- ggplot(rentalOut, aes(x=season2, y=count, fill=season2)) + geom_boxplot()
box5 <- ggplot(rentalOut, aes(x=weekday, y=count, fill=weekday)) + geom_boxplot() + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
box6 <- ggplot(rentalOut, aes(x=workingday, y=count, fill=workingday)) + geom_boxplot()
grid.arrange(box3, box4, box5, box6, ncol=2)
```

### Average Count over Time
```{r}
bar1_data <- rentalOut %>% group_by(weekday) %>% summarise(average_count=sum(count)/n())
bar1 <- ggplot(bar1_data, aes(x=weekday, y=average_count, fill=weekday)) + 
        geom_bar(stat='identity') +
        ggtitle("Average Users by Day") +
        theme(plot.title = element_text(hjust = 0.5), legend.position = "none")

bar2_data <- rentalOut %>% group_by(month) %>% summarise(average_count=sum(count)/n())
bar2 <- ggplot(bar2_data, aes(x=month, y=average_count, fill=month)) + 
        geom_bar(stat='identity') +
        scale_x_discrete(breaks=c("1","2","3","4","5","6","7","8","9","10","11","12"),  
                         labels=c("January","February","March","April","May","June","July",
                                   "August","September","October","November","December")) +
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
        ggtitle("Average Users by Month") +
        theme(plot.title = element_text(hjust = 0.5), legend.position = "none")
grid.arrange(bar1, bar2, nrow=2)
```

There appears to be more rental bike users on Friday and Saturday. There are also more users during the summer months, from May to August.

```{r}
line1_data <- rentalOut %>% group_by(hour,season2) %>% summarise(average_count=sum(count)/n())
line1 <- ggplot(line1_data, aes(x=hour, y=average_count, group=season2, color=season2)) + 
         geom_point() + 
         geom_line() +
         scale_color_discrete(labels = c("spring", "summer", "autumn","winter")) +
         ggtitle("Average Users by Hour for each Season") +
         theme(plot.title = element_text(hjust = 0.5))

line2_data <- rentalOut %>% group_by(hour,weather) %>% summarise(average_count=sum(count)/n())
line2 <- ggplot(line2_data, aes(x=hour, y=average_count, group=weather, color=weather)) + 
    geom_point() + 
    geom_line() +
    scale_color_discrete(labels = c("good", "average", "bad","very bad")) +
    ggtitle("Average Users by Hour for each Weather") +
    theme(plot.title = element_text(hjust = 0.5))
grid.arrange(line1, line2, nrow=2)
```

There are more users during the summer and less users during the winter. The most frequent bike rental timings are at 8am and 5pm, possibly due to working and schooling commuters who use a bike to travel. As expected, there are more users during good weather and less users during bad weather.

```{r}
line3_data <- rentalOut %>% group_by(hour, weekday) %>% summarise(average_count=sum(count)/n())
line3 <- ggplot(line3_data, aes(x=hour, y=average_count, group=weekday, color=weekday)) + 
         geom_point() + 
         geom_line() +
         ggtitle("Average Users by Hour for each Day") +
         theme(plot.title = element_text(hjust = 0.5))

line4_data <- rentalOut %>% group_by(hour) %>% summarise(casual=sum(casual)/n(), registered=sum(registered)/n()) 
line4 <- ggplot(line4_data, aes(x=hour, group=1)) + 
         geom_point(aes(y=casual, color="casual")) + 
         geom_line(aes(y=casual, color="casual")) + 
         geom_point(aes(y=registered, color="registered")) + 
         geom_line(aes(y=registered, color="registered")) +
         ylab("average_count") + 
         labs(color = 'user type') +
         ggtitle("Average Users by Hour for each User Type") +
         theme(plot.title = element_text(hjust = 0.5))

grid.arrange(line3, line4, nrow=2)
```

The bike rental counts are similar during the weekdays but different during the weekends. On the non-working days (weekends), the peak bike rental timing is from 11am to 4pm, which is different from the peak timings of 8am and 5pm on working days. As mentioned earlier, this is most likely due to working and schooling commuters during the weekdays and them not having to travel to work or school on the weekends.

Based on the user type, we can tell that most of the working and schooling commuters are registered users as the graphs seem similar.

### Correlation Analysis

Plot a correlation plot to check the correlation betweeen continuous variables.
```{r}
colPlot <- rentalOut %>% select(temp, atemp, humidity, windspeed, casual, registered, count)
corrplot(cor(colPlot), type="upper", method="color",addCoef.col = "black",number.cex = 0.6)
```

Notice that `temp` and `atemp` are very correlated so we should not include `atemp` in the model. The column `count` is also dependent on `registered` and `casual` so we should not remove these columns as well. 

## Data Splitting

Remove the unnecessary columns, including `holiday` since `workingday` is related to it.
```{r}
rentalKeep <- subset(rentalOut, select = -c(atemp,registered,casual,holiday))
str(rentalKeep)
```

Split the data into training and test sets.
```{r}
set.seed(100)
trainIndex <- createDataPartition(rentalKeep$count, p = 0.8, list=FALSE)
rentalTrain <- rentalKeep[trainIndex,]
rentalTest <- rentalKeep[-trainIndex,]
```

## Modelling

### Linear Regression

We will use linear regression to fit the variables into the data, which uses the ordinary least squares approach to choose the best fit. For variable selection, we will use stepwise regression. The purpose of this algorithm is to add and remove potential candidates in the models and keep those who have a significant impact on the dependent variable. We have chosen the backwards direction so the model will remove insignificant predictors one by one.
```{r}
fit <- lm(count~., data=rentalTrain)

fit1 <- stepAIC(fit, direction="backward")
```

Notice that we have removed the variables `season2` and `month` from the model.
Now, we use the model to predict the test set.
```{r}
ypred <- predict(fit1, rentalTest)
```

To evaluate the model, we will be using the Root Mean Squared Error (RMSE) and R squared.

**Root Mean Squared Error (RMSE)** - Measure of the average deviation of the estimates from the observed values.

**R squared** - Coefficient of determination, scaled between 0 and 1. Measures the amount of variance of the response variable that can be explained by the predictor variables.

```{r}
results1 <- data.frame(
        RMSE = rmse(rentalTest$count,ypred),
        R_squared = summary(fit1)$r.squared
)
results1
```

Let's check its residual diagnostics to check if the model is a good fit.
```{r}
par(mfrow=c(2,2))
plot(fit1)
```

The residual vs fitted plot shows a curve, meaning that the residuals are not evenly distributed around the center line. This indicates that the model may not be a good fit. Let's check the predictions for any errors.
```{r}
summary(ypred)
summary(rentalTest$count)
```

There appears to be negative values predicted for `count`, which does not make sense. 

### Log Transformation

Let's try a log transformation. Perform a stepwise regression to check which variables should be excluded. Compute its RMSE and R squared as well.
```{r}
log_fit <- lm(log(count)~., data=rentalTrain)

log_fit1 <- stepAIC(log_fit, direction="backward")
ypred <- exp(predict(log_fit1, rentalTest))

results2 <- data.frame(
    RMSE = rmse(rentalTest$count,ypred),
    R_squared = summary(log_fit1)$r.squared
)
results2
```

Check its residual diagnostics to check if the model is a good fit.
```{r}
par(mfrow=c(2,2))
plot(log_fit1)
```

The residual vs fitted plot now shows a horizontal line, meaning that the residuals are evenly distributed around the center line. This indicates that the model is a good fit. Let's check the predictions again.
```{r}
summary(ypred)
summary(rentalTest$count)
```

This time the predictions are all positive and appear close to the actual values.

## Results

Plot a barchart to visualise the results of both models.
```{r}
combined <- rbind(results1,results2)
combined$model <- c("Linear", "Log")

bar1 <- ggplot(combined, aes(x=model, y=RMSE, fill=model)) + geom_bar(stat="identity")
bar2 <- ggplot(combined, aes(x=model, y=R_squared, fill=model)) + geom_bar(stat="identity")
grid.arrange(bar1, bar2,ncol=2)
```

Although the log transformation model has a slightly higher RMSE than the linear model, the R squared of the log transformation model is much higher (0.80). This indicates that 80% of the variance of `count` can be explained by the predictor variables, which makes it the better model.

